---
title: "Running BANKSY with Seurat"
date: 'Compiled: `r format(Sys.Date(), "%B %d, %Y")`'
output:
  github_document:
    html_preview: true
    toc: true
  html_document:
    df_print: kable
    theme: simplex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE
)
```

```{r include=FALSE}
t0 <- Sys.time()
```

## Introduction

In this vignette, we describe how to run BANKSY with Seurat objects. If you use 
BANKSY in your research, please cite

> *BANKSY: A Spatial Omics Algorithm that Unifies Cell Type Clustering and Tissue Domain Segmentation*
>
> Vipul Singhal, Nigel Chou, Joseph Lee, Jinyue Liu, Wan Kee Chock, Li Lin, 
Yun-Ching Chang, Erica Teo, Hwee Kuan Lee, Kok Hao Chen, Shyam Prabhakar
>
> bioRxiv, 2022
>
> doi: [10.1101/2022.04.14.488259](https://www.biorxiv.org/content/10.1101/2022.04.14.488259)
>
> Website: https://prabhakarlab.github.io/Banksy


BANKSY is a method that incorporates neighborhood information for clustering 
spatial omics data. By doing so, BANKSY is able to

- improve cell-type assignment in noisy data
- distinguish subtly different cell-types stratified by microenvironment
- identify spatial domains sharing the same microenvironment

The amount of neighborhood information incorporated is controlled by a parameter
`lambda` in [0,1], with higher values giving more weight to the neighbourhood 
information during clustering.

## Overview

The `RunBanksy` function implemented with the *SeuratWrappers* package allows
users to run BANKSY with Seurat objects. We describe two options of running
`RunBanksy`. The first is within Seurat's spatial framework (see 
[here](https://satijalab.org/seurat/articles/spatial_vignette.html) and 
[here](https://satijalab.org/seurat/articles/spatial_vignette_2.html)) and 
requires a Seurat object and a lambda parameter as mandatory input. The second 
option works with Seurat objects that do not have spatial information stored 
within, and therefore requires an additional argument giving the locations of 
the cell centroids or spots. 

Prerequisites to install:

* [Seurat](https://satijalab.org/seurat/install)
* [SeuratData](https://github.com/satijalab/seurat-data)
* [SeuratWrappers](https://github.com/satijalab/seurat-wrappers)
* [Banksy](https://github.com/prabhakarlab/Banksy/)


```{r lib_load}
library(Banksy)
library(Seurat)
library(SeuratData)
library(SeuratWrappers)

library(ggplot2)
library(gridExtra)
library(pals)

# Kelly palette for visualization
mypal <- kelly()[-1]
```

## Running BANKSY within Seurat's spatial framework

We demonstrate how to run BANKSY within Seurat's spatial analysis framework
with a mouse hippocampus Slide-seq v2 dataset from the *SeuratData* package.

After installing *SeuratData*, the data can be accessed as follows:

```{r ss_install, eval = FALSE}
InstallData('ssHippo')
ss.hippo <- LoadData("ssHippo")
``` 

```{r, include=FALSE}
ss.hippo <- LoadData("ssHippo")
``` 

We perform simple preprocessing by filtering beads with high mito percentage and
keeping only beads within the 5th and 98th percentile of total UMI counts. To
keep runtime of this vignette short, we downsample the data to 10,000 beads.

```{r ss_load}
# Filtering
ss.hippo[['percent.mt']] <- PercentageFeatureSet(ss.hippo, pattern = '^MT-')
ss.hippo <- subset(ss.hippo, percent.mt < 10 &
                    nCount_Spatial > quantile(ss.hippo$nCount_Spatial, 0.05) &
                    nCount_Spatial < quantile(ss.hippo$nCount_Spatial, 0.98))
# Downsample
set.seed(42)
ss.hippo <- ss.hippo[,sample(colnames(ss.hippo), 1e4)]
```

Next, normalize the data and find variable genes:

```{r ss_norm}
# Normalize
ss.hippo <- NormalizeData(ss.hippo)
ss.hippo <- FindVariableFeatures(ss.hippo)
```

To run BANKSY, we specify the following:

- `lambda`: a numeric value in [0,1]. With low values of lambda, BANKSY operates
in cell-typing mode, while high values of lambda find spatial domains.
- `assay` and `slot`: determines where to pull the expression data from
- `features`: specifies features for downstream analysis. This can be `'all'`, 
  `'variable'` or a subset of features.  
- `k_geom`: the number of neighbors that defines a cell's neighborhood

Call `?RunBanksy` for more details on function parameters.

```{r ss_banksy}
# Run BANKSY
ss.hippo <- RunBanksy(ss.hippo, lambda = 0.15, verbose=TRUE, 
                      assay = 'Spatial', slot = 'data', features = 'variable',
                      k_geom = 10)
ss.hippo
```

Note that the `RunBanksy` function sets the default assay to `BANKSY` (
determined by the `assay_name` argument). 

The rest of the pipeline is similar to the 'default' Seurat pipline. We 
scale the data and run dimensionality reduction with PCA and UMAP:

```{r ss_pca}
# Scale
ss.hippo <- ScaleData(ss.hippo)
# Run PCA and UMAP
ss.hippo <- RunPCA(ss.hippo, assay = 'BANKSY', features = rownames(ss.hippo), npcs = 50)
ss.hippo <- RunUMAP(ss.hippo, dims = 1:30)
```

Next, find BANKSY clusters:

```{r ss_cluster}
# Clustering
ss.hippo <- FindNeighbors(ss.hippo, dims = 1:30)
ss.hippo <- FindClusters(ss.hippo, resolution = 0.6)
```

Visualize the UMAP and Spatial plot:

```{r ss_viz, fig.align='center', fig.height=5, fig.width=10}
# Viz
grid.arrange(
    DimPlot(ss.hippo, cols = mypal, pt.size = 0.25),
    SpatialDimPlot(ss.hippo, stroke = NA, label = TRUE, label.size = 3, 
                   repel = TRUE, alpha = 0.5, cols = mypal, pt.size.factor = 3),
    ncol = 2
)
```

Find markers based on the BANKSY clusters and visualize them. Here, we find 
differentially expressed genes between the CA1 and CA3 regions.

```{r ss_markers, fig.align='center', fig.height=5, fig.width=10}
# Find markers
DefaultAssay(ss.hippo) <- 'Spatial'
markers <- FindMarkers(ss.hippo, ident.1 = 4, ident.2 = 5, only.pos = F, 
                       logfc.threshold = 1, min.pct = 0.5)
markers <- markers[markers$p_val_adj < 0.01,]
markers
genes <- c('ATP2B1', 'CHGB')
SpatialFeaturePlot(ss.hippo, features = genes, pt.size.factor = 3, 
                   stroke = NA, alpha = 0.5, max.cutoff = 'q95')
```

## Running BANKSY with locations provided explicitly

One can also call `RunBanksy` on a Seurat object created from counts by 
providing the location of cell centroids or spots explicitly. In this case, 
the locations must be stored as metadata. Here, we use a mouse hippocampus 
VeraFISH dataset provided with the *Banksy* package.

```{r hippo_load}
data(hippocampus)
head(hippocampus$expression[,1:5])
head(hippocampus$locations)
```

Construct the Seurat object by storing the locations of cell centroids as 
metadata. We keep cells with total count between 5th and 98th percentile:

```{r hippo_create}
# Create manually
vf.hippo <- CreateSeuratObject(counts = hippocampus$expression,
                               meta.data = hippocampus$locations)
vf.hippo <- subset(vf.hippo,
                   nCount_RNA > quantile(vf.hippo$nCount_RNA, 0.05) & 
                   nCount_RNA < quantile(vf.hippo$nCount_RNA, 0.98))
``` 

Next, we normalize the data by library size and scale the data:

```{r hippo_norm}
# Normalize
vf.hippo <- NormalizeData(vf.hippo, scale.factor = 100, normalization.method = 'RC')
vf.hippo <- ScaleData(vf.hippo)
```

Now, run BANKSY. Here, we provide the column names of the x and y spatial 
coordinates as stored in the metadata to `dimx` and `dimy` respectively:

```{r hippo_banksy}
# Run BANKSY
vf.hippo <- RunBanksy(vf.hippo, lambda = 0.3, dimx = 'sdimx', dimy = 'sdimy', 
                      assay = 'RNA', slot = 'data', features = 'all', k_geom = 10)
```

Scale the BANKSY matrix and run PCA:

```{r hippo_pca}
# Scale
vf.hippo <- ScaleData(vf.hippo)
# PCA
vf.hippo <- RunPCA(vf.hippo, assay = 'BANKSY', features = rownames(vf.hippo), npcs = 50)
```

Find BANKSY clusters:

```{r hippo_cluster}
# Cluster
vf.hippo <- FindNeighbors(vf.hippo, dims = 1:30)
vf.hippo <- FindClusters(vf.hippo, resolution = 0.5)
```

Visualise BANKSY clusters in spatial dimensions:

```{r hippo_viz, fig.align='center', fig.height=5, fig.width=6}
# Viz
FeatureScatter(vf.hippo, 'sdimx', 'sdimy', cols = mypal, pt.size = 0.75)
```

Find markers and visualise them. Here, we do so for a cluster defined by a thin 
layer of cells expressing Gfap. We also write a simple function `genePlot` that 
plots marker genes in spatial dimensions.

```{r hippo_gene, fig.align='center', fig.height=5, fig.width=6}
# Find markers
DefaultAssay(vf.hippo) <- 'RNA'
markers <- FindMarkers(vf.hippo, ident.1 = 6, only.pos = TRUE)

genePlot <- function(object, dimx, dimy, gene, 
                     slot = 'scale.data', q.low = 0.01, q.high = 0.99,
                     col.low='blue', col.high='red') {
    val <- GetAssayData(object, slot)[gene,]
    val.low <- quantile(val, q.low)
    val.high <- quantile(val, q.high)
    val[val < val.low] <- val.low
    val[val > val.high] <- val.high
    pdf <- data.frame(x=object[[dimx]], y=object[[dimy]], gene=val)
    colnames(pdf) <- c('sdimx','sdimy', 'gene')
    ggplot(pdf, aes(x=sdimx,y=sdimy,color=gene)) + geom_point(size = 1) + 
        theme_minimal() + theme(legend.title = element_blank()) +
        scale_color_gradient2(low = col.low, high = col.high) +
        ggtitle(gene)
}

genePlot(vf.hippo, 'sdimx', 'sdimy', 'Gfap')
```

## Getting help

For more information, visit https://github.com/prabhakarlab/Banksy. 

<details>
  <summary>Vignette runtime</summary>

```{r echo=FALSE}
Sys.time() - t0
```

</details>


<details>
  <summary>Session info</summary>
  
```{r, sess}
sessionInfo()
```

</details>
