---
title: "Running BANKSY with Seurat"
date: 'Compiled: `r format(Sys.Date(), "%B %d, %Y")`'
output:
  github_document:
    html_preview: true
    toc: true
  html_document:
    df_print: kable
    theme: simplex
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE
)
```

```{r include=FALSE}
t0 <- Sys.time()
```

## Introduction

In this vignette, we describe how to run BANKSY with Seurat objects. If you use 
BANKSY in your research, please cite

> *BANKSY unifies cell typing and tissue domain segmentation for scalable spatial omics data analysis*
>
> Vipul Singhal, Nigel Chou, Joseph Lee, Yifei Yue, Jinyue Liu, Wan Kee Chock, Li Lin, Yun-Ching Chang, Erica Mei Ling Teo, Jonathan Aow, Hwee Kuan Lee, Kok Hao Chen & Shyam Prabhakar
>
> Nature Genetics, 2024
>
> doi: [10.1038/s41588-024-01664-3](https://doi.org/10.1038/s41588-024-01664-3)
> 
> Website: https://prabhakarlab.github.io/Banksy

BANKSY is a method that incorporates neighborhood information for clustering 
spatial omics data. By doing so, BANKSY is able to

- improve cell-type assignment in noisy data
- distinguish subtly different cell-types stratified by microenvironment
- identify spatial domains sharing the same microenvironment

The amount of neighborhood information incorporated is controlled by a parameter
`lambda` in [0,1], with higher values giving more weight to the neighbourhood 
information during clustering.

## Overview

The `RunBanksy` function implemented with the *SeuratWrappers* package allows
users to run BANKSY with Seurat objects. We describe two options of running
`RunBanksy`. The first is within Seurat's spatial framework (see 
[here](https://satijalab.org/seurat/articles/spatial_vignette.html) and 
[here](https://satijalab.org/seurat/articles/spatial_vignette_2.html)) and 
requires a Seurat object and a lambda parameter as mandatory input. The second 
option works with Seurat objects that do not have spatial information stored 
within, and therefore requires an additional argument giving the locations of 
the cell centroids or spots. 

**Caveat**: `ScaleData` should not be run after a call to `RunBanksy`; `RunBanksy`
populates the `scale.data` slot with the scaled BANKSY matrix. Calling `ScaleData`
after `RunBanksy` performs gene-wise z-scaling, negating the effect of `lambda`. 

Prerequisites to install:

* [Seurat](https://satijalab.org/seurat/install)
* [SeuratData](https://github.com/satijalab/seurat-data)
* [SeuratWrappers](https://github.com/satijalab/seurat-wrappers)
* [Banksy](https://github.com/prabhakarlab/Banksy/)


```{r lib_load}
library(Banksy)
library(Seurat)
library(SeuratData)
library(SeuratWrappers)

library(ggplot2)
library(gridExtra)
library(pals)

# Kelly palette for visualization
mypal <- kelly()[-1]
```

## Running BANKSY within Seurat's spatial framework

We demonstrate how to run BANKSY within Seurat's spatial analysis framework
with a mouse hippocampus Slide-seq v2 dataset from the *SeuratData* package.

After installing *SeuratData*, the data can be accessed as follows:

```{r ss_install, eval = FALSE}
InstallData('ssHippo')
ss.hippo <- LoadData("ssHippo")
``` 

```{r, include=FALSE}
ss.hippo <- LoadData("ssHippo")
``` 

We perform simple preprocessing by filtering beads with high mito percentage and
keeping only beads within the 5th and 98th percentile of total UMI counts. To
keep runtime of this vignette short, we downsample the data to 10,000 beads.

```{r ss_load}
# Filtering
ss.hippo[['percent.mt']] <- PercentageFeatureSet(ss.hippo, pattern = '^MT-')
ss.hippo <- subset(ss.hippo, percent.mt < 10 &
                    nCount_Spatial > quantile(ss.hippo$nCount_Spatial, 0.05) &
                    nCount_Spatial < quantile(ss.hippo$nCount_Spatial, 0.98))
# Downsample
set.seed(42)
ss.hippo <- ss.hippo[,sample(colnames(ss.hippo), 1e4)]
```

Next, normalize the data and find variable genes:

```{r ss_norm}
# Normalize
ss.hippo <- NormalizeData(ss.hippo)
ss.hippo <- FindVariableFeatures(ss.hippo)
ss.hippo <- ScaleData(ss.hippo)
```

To run BANKSY, we specify the following:

- `lambda`: a numeric value in [0,1]. With low values of lambda, BANKSY operates
in cell-typing mode, while high values of lambda find spatial domains.
- `assay` and `slot`: determines where to pull the expression data from
- `features`: specifies features for downstream analysis. This can be `'all'`, 
  `'variable'` or a subset of features.  
- `k_geom`: the number of neighbors that defines a cell's neighborhood

Call `?RunBanksy` for more details on function parameters.

```{r ss_banksy}
# Run BANKSY
ss.hippo <- RunBanksy(ss.hippo, lambda = 0.2, verbose=TRUE, 
                      assay = 'Spatial', slot = 'data', features = 'variable',
                      k_geom = 15)
ss.hippo
```

Note that the `RunBanksy` function sets the default assay to `BANKSY` (
determined by the `assay_name` argument) and fills the `scale.data` slot. Users
should not call `ScaleData` on the `BANKSY` assay as this negates the effects
of `lambda`. 

The rest of the pipeline is similar to the 'default' Seurat pipline. We 
scale the data and run dimensionality reduction with PCA and UMAP:

```{r ss_pca}
# Run PCA and UMAP
ss.hippo <- RunPCA(ss.hippo, assay = 'BANKSY', features = rownames(ss.hippo), npcs = 30)
ss.hippo <- RunUMAP(ss.hippo, dims = 1:30)
```

Next, find BANKSY clusters:

```{r ss_cluster}
# Clustering
ss.hippo <- FindNeighbors(ss.hippo, dims = 1:30)
ss.hippo <- FindClusters(ss.hippo, resolution = 0.5)
```

Visualize the UMAP and Spatial plot:

```{r ss_viz, fig.align='center', fig.height=5, fig.width=10}
# Viz
coords <- GetTissueCoordinates(ss.hippo)
coords$cluster <- Idents(ss.hippo)[coords$cells]
spatial_plot <- ggplot(coords, aes(x = x, y = y, color = cluster)) +
    geom_point(size = 0.25) + theme_minimal() +
    guides(color = guide_legend(override.aes = list(size = 3)))
grid.arrange(
    DimPlot(ss.hippo, pt.size = 0.25, label = TRUE, label.size = 3, repel = TRUE),
    spatial_plot,
    ncol = 2
)
```

Find markers based on the BANKSY clusters and visualize them. Here, we find 
differentially expressed genes between the CA1 and CA3 regions.

```{r ss_markers, fig.align='center', fig.height=5, fig.width=10}
# Find markers
DefaultAssay(ss.hippo) <- 'Spatial'
markers <- FindMarkers(ss.hippo, ident.1 = 4, ident.2 = 9, only.pos = F,
                       logfc.threshold = 1, min.pct = 0.5)
markers <- markers[markers$p_val_adj < 0.01,]
markers
genes <- c('ATP2B1', 'CHGB')
coords <- GetTissueCoordinates(ss.hippo)
gene_plots <- lapply(genes, function(g) {
    val <- GetAssayData(ss.hippo, assay = 'Spatial', layer = 'data')[g, coords$cells]
    val[val > quantile(val, 0.95)] <- quantile(val, 0.95)
    coords$expr <- val
    ggplot(coords, aes(x = x, y = y, color = expr)) +
        geom_point(size = 0.25) + theme_minimal() +
        scale_color_gradient(low = 'grey90', high = 'red') +
        ggtitle(g) + theme(legend.title = element_blank())
})
grid.arrange(grobs = gene_plots, ncol = 2)
```

## Running BANKSY with locations provided explicitly

One can also call `RunBanksy` on a Seurat object created from counts by 
providing the location of cell centroids or spots explicitly. In this case, 
the locations must be stored as metadata. Here, we use a mouse hippocampus 
VeraFISH dataset provided with the *Banksy* package.

```{r hippo_load}
data(hippocampus)
head(hippocampus$expression[,1:5])
head(hippocampus$locations)
```

Construct the Seurat object by storing the locations of cell centroids as 
metadata. We keep cells with total count between 5th and 98th percentile:

```{r hippo_create}
# Create manually
vf.hippo <- CreateSeuratObject(counts = hippocampus$expression,
                               meta.data = hippocampus$locations)
vf.hippo <- subset(vf.hippo,
                   nCount_RNA > quantile(vf.hippo$nCount_RNA, 0.05) & 
                   nCount_RNA < quantile(vf.hippo$nCount_RNA, 0.98))
``` 

Next, we normalize the data by library size and scale the data:

```{r hippo_norm}
# Normalize
vf.hippo <- NormalizeData(vf.hippo, scale.factor = 100, normalization.method = 'RC')
vf.hippo <- ScaleData(vf.hippo)
```

Now, run BANKSY. Here, we provide the column names of the x and y spatial 
coordinates as stored in the metadata to `dimx` and `dimy` respectively:

```{r hippo_banksy}
# Run BANKSY
vf.hippo <- RunBanksy(vf.hippo, lambda = 0.2, dimx = 'sdimx', dimy = 'sdimy', 
                      assay = 'RNA', slot = 'data', features = 'all', k_geom = 10)
```

Note that the `RunBanksy` function sets the default assay to `BANKSY` (
determined by the `assay_name` argument) and fills the `scale.data` slot. Users
should not call `ScaleData` on the `BANKSY` assay as this negates the effects
of `lambda`. 

Run PCA on the BANKSY matrix:

```{r hippo_pca}
# PCA
vf.hippo <- RunPCA(vf.hippo, assay = 'BANKSY', features = rownames(vf.hippo), npcs = 20)
```

Find BANKSY clusters:

```{r hippo_cluster}
# Cluster
vf.hippo <- FindNeighbors(vf.hippo, dims = 1:20)
vf.hippo <- FindClusters(vf.hippo, resolution = 0.5)
```

Visualise BANKSY clusters in spatial dimensions:

```{r hippo_viz, fig.align='center', fig.height=6, fig.width=7}
# Viz
FeatureScatter(vf.hippo, 'sdimx', 'sdimy', cols = mypal, pt.size = 0.75)
FeatureScatter(vf.hippo, 'sdimx', 'sdimy', cols = mypal, pt.size = 0.1) + facet_wrap(~ colors)
```

Find markers and visualise them. Here, we do so for a cluster defined by a thin 
layer of cells expressing Gfap. We also write a simple function `genePlot` that 
plots marker genes in spatial dimensions.

```{r hippo_gene, fig.align='center', fig.height=5, fig.width=6}
# Find markers
DefaultAssay(vf.hippo) <- 'RNA'
markers <- FindMarkers(vf.hippo, ident.1 = 6, only.pos = TRUE)

genePlot <- function(object, dimx, dimy, gene, assay = 'RNA',
                     layer = 'scale.data', q.low = 0.01, q.high = 0.99,
                     col.low='blue', col.high='red') {
    val <- GetAssayData(object, assay=assay, layer=layer)[gene,]
    val.low <- quantile(val, q.low)
    val.high <- quantile(val, q.high)
    val[val < val.low] <- val.low
    val[val > val.high] <- val.high
    pdf <- data.frame(x=object[[dimx]], y=object[[dimy]], gene=val)
    colnames(pdf) <- c('sdimx','sdimy', 'gene')
    ggplot(pdf, aes(x=sdimx,y=sdimy,color=gene)) + geom_point(size = 1) + 
        theme_minimal() + theme(legend.title = element_blank()) +
        scale_color_gradient2(low = col.low, high = col.high) +
        ggtitle(gene)
}

genePlot(vf.hippo, 'sdimx', 'sdimy', 'Gfap')
```

## Multi-sample analysis

This section demonstrate demonstrates multi-sample analysis. Such an approach
is appropriate when analysing multiple spatial omics datasets with 
non-contiguous spatial coordinates, and when large batch effects are not 
present. 

Here, we use a mouse hippocampus 
VeraFISH dataset provided with the *Banksy* package.

```{r}
data(hippocampus)
head(hippocampus$expression[,1:5])
head(hippocampus$locations)
```

For demonstration purposes, we create three separate datasets by splitting the
data. 

```{r}
# Number of groups
n_groups = 3
group_names = paste0('group', seq(n_groups))
group_size = 1000
starts = seq(1, by=group_size, length.out=n_groups)
ends = starts + group_size - 1

# List of Seurat objects
seu_list = lapply(seq(n_groups), function(i) {
  idx = seq(starts[i], ends[i])
  seu = CreateSeuratObject(
    counts = hippocampus$expression[,idx],
    meta.data = data.frame(scale(hippocampus$locations[idx,], scale = FALSE))
  )
  # Set original identity of cell
  seu$orig.ident = group_names[i]
  seu
})
seu_list
```

Perform normalisation for each dataset. 

```{r}
seu_list = lapply(seu_list, NormalizeData,
                  scale.factor = 100, normalization.method = 'RC')
```

Merge the datasets. Note that the spatial coordinates overlap.

```{r multi-spatial, fig.align='center', fig.height=6, fig.width=9}
# Merge
seu = Reduce(merge, seu_list)
seu = JoinLayers(seu) # run this for Seurat v5 objects

# Plot spatial coordinates colored by group
plot(FetchData(seu, c('sdimx', 'sdimy')), col = factor(seu$orig.ident))
```

Now run BANKSY. For multi-sample analysis, the argument `group` must be 
provided, which specifies the name of the metadata column that gives the 
assignment of each cell or spot to its original Seurat object. Here, we use
`orig.ident`. Internally, providing the `group` argument tells the function to
compute neighborhood matrices based on locations staggered by `group`, 
ensuring that cells from different spatial datasets do not overlap. The 
staggered locations are stored in the metadata for sanity checking. The 
`split.scale` argument allows for within-group scaling, accounting for minor 
differences in datasets.

```{r}
# Grouping variable
head(seu@meta.data)
table(seu$orig.ident)

# Run BANKSY
seu = RunBanksy(seu, lambda = 0.2, assay = 'RNA', slot = 'data',
                dimx = 'sdimx', dimy = 'sdimy', features = 'all',
                group = 'orig.ident', split.scale = TRUE, k_geom = 15)

# Staggered locations added to metadata
head(seu@meta.data)
```

The rest of the workflow follows as before:

```{r}
seu = RunPCA(seu, assay = 'BANKSY', features = rownames(seu), npcs = 30)
seu = RunUMAP(seu, dims = 1:30)
seu = FindNeighbors(seu, dims = 1:30)
seu = FindClusters(seu, resolution = 1)
```

Visualise clusters:

```{r multi-umap, fig.align='center', fig.height=5, fig.width=5}
mypal <- kelly()[-1]
DimPlot(seu, pt.size = 0.25, label = TRUE, label.size = 3, cols = mypal)
```

```{r multi-spatial-staggered, fig.align='center', fig.height=3, fig.width=9}
FeatureScatter(seu, 'staggered_sdimx', 'staggered_sdimy', pt.size = 0.75, cols = mypal)
```

## Spatial data integration with Harmony

BANKSY can be used with Harmony for integrating multiple  spatial omics 
datasets in the presence of strong batch effects. 

Download the data. 

```{r harmony_download}
library(spatialLIBD)
library(ExperimentHub)
library(harmony)

ehub <- ExperimentHub::ExperimentHub()
spe <- spatialLIBD::fetch_data(type = "spe", eh = ehub)

imgData(spe) <- NULL
assay(spe, "logcounts") <- NULL
reducedDims(spe) <- NULL
rowData(spe) <- NULL
colData(spe) <- DataFrame(
  sample_id = spe$sample_id,
  clust_annotation = factor(
    addNA(spe$layer_guess_reordered_short),
    exclude = NULL, labels = seq(8)
  ),
  in_tissue = spe$in_tissue,
  row.names = colnames(spe)
)
invisible(gc())

# Subset to first sample of each subject
sample_names <- c("151507", "151669", "151673")
spe_list <- lapply(sample_names, function(x) spe[, spe$sample_id == x])
rm(spe)
invisible(gc())
```

Normalise the data and compute highly variable features. 

```{r harmony_norm}
# Convert to Seurat and Normalize data
seu_list <- lapply(spe_list, function(x) {
  x <- as.Seurat(x, data = NULL)
  NormalizeData(x, scale.factor = 3000, normalization.method = 'RC')
})

# Compute HVGs for each dataset and take the union
hvgs <- lapply(seu_list, function(x) {
  VariableFeatures(FindVariableFeatures(x, nfeatures = 2000))
})
hvgs <- Reduce(union, hvgs)

# Subset to HVGs
seu_list <- lapply(seu_list, function(x) x[hvgs,])
seu <- Reduce(merge, seu_list)

locs <- do.call(rbind.data.frame, lapply(spe_list, spatialCoords))
seu@meta.data <- cbind(seu@meta.data, locs)

seu
```

Run BANKSY. When analysing multiple samples, the argument `group` must be 
provided, which specifies the name of the metadata column that gives the 
assignment of each cell or spot to its original Seurat object. Here, we use
`sample_id`. Internally, providing the `group` argument tells the function to
compute neighborhood matrices based on locations staggered by `group`, 
ensuring that cells from different spatial datasets do not overlap. The 
staggered locations are stored in the metadata for sanity checking. 
Within-group scaling has little effect in the presence of strong batch effects, 
hence, we set `split.scale=FALSE` for efficiency. 

```{r harmony_banksy}
# Grouping variable
head(seu@meta.data)
table(seu$sample_id)

sdimx <- 'pxl_col_in_fullres'
sdimy <- 'pxl_row_in_fullres'

# Run BANKSY
seu <- RunBanksy(seu, lambda = 0.2, assay = 'originalexp', slot = 'data',
                dimx = sdimx, dimy = sdimy, features = 'all',
                group = 'sample_id', split.scale = FALSE, k_geom = 6)
```

Compute a spatially-aware embedding with PCA on the BANKSY matrix, and run 
Harmony on this embedding.

```{r harmony_pca}
seu <- RunPCA(seu, assay = 'BANKSY', features = rownames(seu), npcs = 10)
seu <- RunHarmony(seu, group.by.vars='sample_id')
```

The rest of the workflow follows as before:

```{r harmony_cluster}
seu <- RunUMAP(seu, dims = 1:10, reduction = 'harmony')
seu <- FindNeighbors(seu, dims = 1:10, reduction = 'harmony')
seu <- FindClusters(seu, resolution = 0.4)
```

Visualise clusters:

```{r harmony_viz, fig.align='center', fig.height=5, fig.width=10}
grid.arrange(
    DimPlot(seu, pt.size = 0.25, label = TRUE, label.size = 3, cols = mypal),
    DimPlot(seu, pt.size = 0.25, group.by = 'sample_id'),
    ncol = 2
)
```

```{r harmony_spatial, fig.align='center', fig.height=4, fig.width=12}
spatial_plots <- lapply(sample_names, function(s) {
    sub <- subset(seu, sample_id == s)
    ggplot(sub@meta.data, aes(x = pxl_col_in_fullres, y = pxl_row_in_fullres,
                              color = seurat_clusters)) +
        geom_point(size = 1) + coord_equal() + theme_minimal() +
        scale_color_manual(values = mypal) +
        ggtitle(s) + theme(legend.position = 'none') +
        labs(x = NULL, y = NULL)
})
grid.arrange(grobs = spatial_plots, ncol = 3)
```

## Scaling to large datasets

For large datasets (millions of cells), materializing the full BANKSY matrix can
exceed memory limits. The `compute_pca` option in `RunBanksy` avoids this by
computing PCA directly via a lazy linear operator. The BANKSY matrix is never
formed; instead, matrix-vector products are evaluated on-the-fly using the sparse
expression matrix and weight matrix. This reduces peak memory from
$O(\text{genes} \times \text{cells})$ (the dense BANKSY matrix) to
$O(\text{nnz}(\text{input}))$ (the number of non-zero entries in the sparse
input), since only the sparse input and sparse weight matrix are held in memory.

**Note**: `compute_pca=TRUE` currently supports M=0 only (no azimuthal Gabor
filter).

This feature requires the development versions of Banksy and SeuratWrappers:

```{r install_dev, eval=FALSE}
devtools::install_github('prabhakarlab/Banksy@feat-sparse-matmul')
devtools::install_github('jleechung/seurat-wrappers@feat-sparse-matmul')
```

We demonstrate this with a 10X Xenium mouse brain dataset (~37k cells),
performing domain segmentation with `lambda = 0.8`.

Download the dataset:

```{r xenium_download, eval=FALSE}
download.file(
    url = 'https://cf.10xgenomics.com/samples/xenium/1.0.2/Xenium_V1_FF_Mouse_Brain_Coronal_Subset_CTX_HP/Xenium_V1_FF_Mouse_Brain_Coronal_Subset_CTX_HP_outs.zip',
    destfile = '~/Downloads/xenium_subset.zip'
)
unzip('~/Downloads/xenium_subset.zip', exdir = '~/Downloads/xenium_subset')
```

Load and preprocess:

```{r xenium_load}
xenium <- LoadXenium('~/Downloads/xenium_subset', fov = 'fov')
xenium <- NormalizeData(xenium)
xenium <- FindVariableFeatures(xenium)
```

With `compute_pca = TRUE`, `RunBanksy` computes the BANKSY PCA embedding
directly and stores it as a dimensionality reduction. No `BANKSY` assay is
created, and there is no need to call `RunPCA` separately:

```{r xenium_lazy}
xenium_input <- xenium
xenium <- RunBanksy(xenium, lambda = 0.8, verbose = TRUE,
                    assay = 'Xenium', slot = 'data', features = 'variable',
                    use_agf = FALSE, k_geom = 30,
                    assay_name = 'BANKSY',
                    compute_pca = TRUE, npcs = 50)
xenium <- FindNeighbors(xenium, reduction = 'BANKSY', dims = 1:50)
xenium <- FindClusters(xenium, resolution = 0.5)
```

Visualize the spatial domains:

```{r xenium_viz, fig.align='center', fig.height=5, fig.width=7}
ImageDimPlot(xenium, size = 0.5)
```

<details>
<summary>Equivalence with the standard workflow</summary>

The standard workflow constructs the full BANKSY matrix
$$\mathbf{M} = \begin{bmatrix} \sqrt{1-\lambda}\;\mathbf{Z}(\mathbf{X}) \\[4pt] \sqrt{\lambda}\;\mathbf{Z}(\mathbf{X}\mathbf{W}) \end{bmatrix}$$
where $\mathbf{X}$ is the expression matrix ($g \times n$), $\mathbf{W}$ is the
sparse neighbor-weight matrix ($n \times n$, $k$ non-zeros per column), and
$\mathbf{Z}(\cdot)$ denotes row-wise z-scoring followed by clipping to
$[-10, 10]$ (matching Seurat's `FastRowScale`). PCA is then computed on
$\mathbf{M}$ via `RunPCA`.

With `compute_pca=TRUE`, $\mathbf{M}$ is never formed. Instead, `irlba`
accesses $\mathbf{M}$ through a lazy linear operator that evaluates
matrix--vector products on the fly. Writing the row-wise z-score as
$\mathbf{Z}(\mathbf{A})_{ij} = (A_{ij} - \mu_i) / \sigma_i$, since
`irlba` only requires the ability to compute forward products
$\mathbf{M}\mathbf{v}$ and adjoint products $\mathbf{M}^{\!\top}\mathbf{u}$
for arbitrary vectors $\mathbf{v} \in \mathbb{R}^n$ and
$\mathbf{u} \in \mathbb{R}^{2g}$, rather than access to $\mathbf{M}$ itself,
each block of the forward product is evaluated as:

$$\mathbf{Z}(\mathbf{X})\,\mathbf{v}
= \frac{\mathbf{X}\mathbf{v} - \boldsymbol{\mu}\,\mathbf{1}^{\!\top}\mathbf{v}}
       {\boldsymbol{\sigma}}$$

$$\mathbf{Z}(\mathbf{X}\mathbf{W})\,\mathbf{v}
= \frac{\mathbf{X}(\mathbf{W}\mathbf{v}) - \boldsymbol{\mu}_{H_0}\,\mathbf{1}^{\!\top}\mathbf{v}}
       {\boldsymbol{\sigma}_{H_0}}$$

where the key step is $(\mathbf{X}\mathbf{W})\mathbf{v} = \mathbf{X}(\mathbf{W}\mathbf{v})$
by associativity: the sparse $\mathbf{W}$ is applied to $\mathbf{v}$ first
($O(kn)$), then the result is left-multiplied by the sparse $\mathbf{X}$
($O(\mathrm{nnz}(\mathbf{X}))$), avoiding formation of the dense
$g \times n$ product $\mathbf{X}\mathbf{W}$. The adjoint is derived
analogously. Row means,
standard deviations, and a sparse correction for Seurat's z-score
clipping are precomputed once before the Lanczos iterations begin.

This reduces peak memory from $O(gn)$ (the dense BANKSY matrix) to
$O(\mathrm{nnz}(\mathbf{X}) + kn)$ (the sparse input plus the sparse
weight matrix), while producing numerically identical PCA embeddings.

The standard workflow materializes the full BANKSY matrix as a Seurat assay,
then runs PCA via `RunPCA`:

```{r xenium_standard}
xenium_std <- RunBanksy(xenium_input, lambda = 0.8, verbose = TRUE,
                        assay = 'Xenium', slot = 'data', features = 'variable',
                        use_agf = FALSE, k_geom = 30)
xenium_std <- RunPCA(xenium_std, assay = 'BANKSY', features = rownames(xenium_std),
                     npcs = 50)
xenium_std <- FindNeighbors(xenium_std, dims = 1:50)
xenium_std <- FindClusters(xenium_std, resolution = 0.5)
```

```{r xenium_std_viz, fig.align='center', fig.height=5, fig.width=7}
ImageDimPlot(xenium_std, size = 0.5)
```

We compare the two embeddings with three checks:

```{r xenium_check, fig.align='center', fig.height=4, fig.width=6}
U1 <- scale(Embeddings(xenium_std, 'pca'), scale = FALSE)
U2 <- scale(Embeddings(xenium, 'BANKSY'), scale = FALSE)
npcs <- ncol(U1)
```

**Subspace overlap.** Individual PCs may differ due to near-degenerate singular
values, but the subspaces they span should be equivalent. We measure this via
principal angles between the top-k subspaces from each path. A minimum cosine
near 1 means the two subspaces are effectively identical:

```{r xenium_subspace, fig.align='center', fig.height=4, fig.width=6}
subspace_cos <- sapply(seq_len(npcs), function(k) {
    Q1 <- qr.Q(qr(U1[, 1:k, drop = FALSE]))
    Q2 <- qr.Q(qr(U2[, 1:k, drop = FALSE]))
    min(svd(crossprod(Q1, Q2))$d)
})

plot(seq_len(npcs), subspace_cos, type = 'b', pch = 19, ylim = c(0.5, 1),
     xlab = 'Number of PCs (k)', ylab = 'Min cosine of principal angles',
     main = 'Subspace overlap: standard vs compute_pca')
abline(h = 0.99, lty = 2, col = 'grey50')
```

**kNN overlap.** Since clustering operates on the nearest-neighbor graph, we
compare the 30-NN graphs built from each embedding. The mean Jaccard index
measures what fraction of each cell's neighbors are shared between the two
paths:

```{r xenium_knn}
nn1 <- RANN::nn2(U1, k = 30)$nn.idx
nn2 <- RANN::nn2(U2, k = 30)$nn.idx
jaccard <- mean(sapply(seq_len(nrow(nn1)), function(i) {
    length(intersect(nn1[i,], nn2[i,])) / length(union(nn1[i,], nn2[i,]))
}))
cat('Mean kNN Jaccard overlap (k=30):', round(jaccard, 4), '\n')
```

**Clustering agreement.** The adjusted Rand index (ARI) measures how closely
the two sets of cluster labels agree, adjusted for chance:

```{r xenium_ari}
ari <- mclust::adjustedRandIndex(
    as.integer(Idents(xenium_std)),
    as.integer(Idents(xenium))
)
cat('Adjusted Rand Index:', round(ari, 4), '\n')
```

</details>

## Getting help

For more information, visit https://github.com/prabhakarlab/Banksy. 

<details>
  <summary>Vignette runtime</summary>

```{r echo=FALSE}
Sys.time() - t0
```

</details>


<details>
  <summary>Session info</summary>
  
```{r, sess}
sessionInfo()
```

</details>
