---
title: "Running CIARA on Seurat Objects"
date: 'Compiled: `r format(Sys.Date(), "%B %d, %Y")`'
output: 
  github_document:
    html_preview: true
    toc: true
    toc_depth: 3
    fig_width: 16
  html_document:
    df_print: kable
    theme: united
    fig_height: 5
    fig_width: 16
    out_height: 4
---

This vigettte demonstrates how to run CIARA (Cluster Independent Algorithm for the identification of RAre cell types) on Seurat objects. Functions and parameters are based on the CRAN package [CIARA](https://CRAN.R-project.org/package=CIARA). If you use CIARA, please cite:
> CIARA: Cluster Independent Algorithm for Rare Cell Types Identification
>
> Gabriele Lubatti
>
> CRAN: https://CRAN.R-project.org/package=CIARA



```{r setup, include=FALSE}
knitr::opts_chunk$set(
  tidy = TRUE,
  tidy.opts = list(width.cutoff = 95),
  message = FALSE,
  warning = FALSE
)
options(timeout = 1000)
```

Prerequisites to install: 

* [Seurat](https://satijalab.org/seurat/install)
* [SeuratWrappers](https://github.com/satijalab/seurat-wrappers)
* [SeuratData](https://github.com/satijalab/seurat-data)
* [CIARA](https://CRAN.R-project.org/package=CIARA)

```{r packages}
library(Seurat)
library(SeuratWrappers)
library(SeuratData)
```

## Analysis of human PBMC

To learn more about this dataset, type `?pbmc_small`

```{r}
InstallData("pbmcsca")
data("pbmc_small")
```



```{r}
pbmc_small <- FindVariableFeatures(pbmc_small, selection.method = "vst", nfeatures = 2000)
pbmc_small <- ScaleData(pbmc_small, verbose = FALSE)
pbmc_small <- RunPCA(pbmc_small, npcs = 30, verbose = FALSE)
pbmc_small <- RunUMAP(pbmc_small, reduction = "pca", dims = 1:20, set.seed = 42)
pbmc_small <- FindNeighbors(pbmc_small, reduction = "pca", dims = 1:30, k.param = 5)
pbmc_small <- FindClusters(pbmc_small, resolution = 0.1)
pbmc_small <- get_background_full_seurat(pbmc_small,threshold = 1, n_cells_low = 3, n_cells_high = 20)
    
```



## Run CIARA 

CIARA (Cluster Independent Algorithm for the identification of RAre cell types) is a cluster independent approach that selects genes localized in a small number of neighboring cells from high dimensional PCA space.
We don't execute the CIARA algorithm and we directly load the result

```{r }
pbmc_small <- CIARA_seurat( pbmc_small, cores_number = 1, p_value = 0.001, odds_ratio = 2, local_region = 1, approximation = FALSE)
```

## Identifying highly localized genes 

```{r }
result <- pbmc_small[["RNA"]][['CIARA_p_value']]
genes_name <- row.names(result)[! (is.na(result[,1]))]
result <- result[genes_name, ]
names(result) <- genes_name


ciara_genes <- names(result)[result<1]
ciara_genes_top <- names(result)[order(as.numeric(result))]



```


```{r }
FeaturePlot(pbmc_small, features = ciara_genes_top[1:2])
```
```{r}
DimPlot(pbmc_small)
```
```{r}
DotPlot(pbmc_small, features = ciara_genes_top[1:10]) + RotatedAxis()
```


For more down stream analysis, including cluster analysis using CIARA highly localized genes, please see  utils::vignette("CIARA") in [CIARA](https://CRAN.R-project.org/package=CIARA).


```{r}
utils::sessionInfo()
```


