---
title: "Running Coralysis on Seurat Objects"
date: 'Compiled: `r format(Sys.Date(), "%B %d, %Y")`'
output: 
  html_document:
    df_print: kable
    theme: united
    fig_height: 5
    fig_width: 16
    out_height: 4
  github_document:
    html_preview: true
    toc: true
    toc_depth: 3
    fig_width: 16
---

This vigettte demonstrates how to run Coralysis on Seurat objects. Parameters and commands are based off of the [Coralysis reference manual](https://bioconductor.org/packages/devel/bioc/manuals/Coralysis/man/Coralysis.pdf). If you use Coralysis, please cite:

> *Coralysis enables sensitive identification of imbalanced cell types and states in single-cell data via multi-level integration*
>
> António GG Sousa, Johannes Smolander, Sini Junttila & Laura L Elo 
>
> bioRxiv, 2025
>
> doi: [10.1101/2025.02.07.637023](https://doi.org/10.1101/2025.02.07.637023)
>
> Bioconductor: https://bioconductor.org/packages/devel/bioc/html/Coralysis.html

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  tidy = TRUE,
  tidy.opts = list(width.cutoff = 95),
  message = FALSE,
  warning = FALSE
)
options(timeout = 1000)
```

Prerequisites to install: 

* [Seurat](https://satijalab.org/seurat/install)
* [SeuratWrappers](https://github.com/satijalab/seurat-wrappers)
* [SeuratData](https://github.com/satijalab/seurat-data)

```{r packages}
library(Seurat)
library(SeuratData)
library(SeuratWrappers)
```

## {.tabset .tabset-pills}

### Systematic comparative analysis of human PBMC

To learn more about this dataset, type `?pbmcsca` 

```{r pbmcsca, results='hide'}
InstallData("pbmcsca")
pbmcsca <- LoadData("pbmcsca")
pbmcsca <- UpdateSeuratObject(pbmcsca)
pbmcsca[["RNA"]] <- split(pbmcsca[["RNA"]], f = pbmcsca$Method)
pbmcsca <- NormalizeData(pbmcsca)
pbmcsca <- FindVariableFeatures(pbmcsca)
pbmcsca <- ScaleData(pbmcsca)
pbmcsca <- RunPCA(pbmcsca)
set.seed(1239)
pbmcsca <- IntegrateLayers(object = pbmcsca, 
                           method = CoralysisIntegration, 
                           new.reduction = "integrated.coralysis",
                           batch = "Method",
                           threads = 4)
pbmcsca <- RunUMAP(pbmcsca, reduction = 'integrated.coralysis', dims = 1:30)
pbmcsca <- FindNeighbors(pbmcsca, reduction = 'integrated.coralysis', dims = 1:30)
pbmcsca <- FindClusters(pbmcsca)
DimPlot(pbmcsca, group.by = c('Method', 'ident',"CellType"), ncol = 3)
```

```{r cleanup_pbmcsca, include=FALSE}
rm(pbmcsca)
gc(verbose = FALSE)
```

### Interferon-stimulated and control PBMC

To learn more about this dataset, type `?ifnb` 

```{r ifnb_stim, results='hide'}
InstallData("ifnb")
ifnb <- LoadData("ifnb")
ifnb <- UpdateSeuratObject(ifnb)
ifnb[["RNA"]] <- split(ifnb[["RNA"]], f = ifnb$stim)
ifnb <- NormalizeData(ifnb)
ifnb <- FindVariableFeatures(ifnb)
ifnb <- ScaleData(ifnb)
ifnb <- RunPCA(ifnb)
set.seed(45)
ifnb <- IntegrateLayers(object = ifnb, 
                        method = CoralysisIntegration, 
                        new.reduction = "integrated.coralysis",
                        batch = "stim",
                        threads = 4)
ifnb <- RunUMAP(ifnb, reduction = 'integrated.coralysis', dims = 1:30)
ifnb <- FindNeighbors(ifnb, reduction = 'integrated.coralysis', dims = 1:30)
ifnb <- FindClusters(ifnb)
DimPlot(ifnb, group.by = c('stim', 'ident', 'seurat_annotations'), ncol = 3)
```

```{r cleanup_ifnb, include=FALSE}
rm(ifnb)
gc(verbose = FALSE)
```

### Eight human pancreatic islet datasets

To learn more about this dataset, type `?panc8` 

```{r pancreas, results='hide'}
InstallData("panc8")
panc8 <- LoadData("panc8")
panc8 <- UpdateSeuratObject(panc8)
panc8[["RNA"]] <- split(panc8[["RNA"]], f = panc8$replicate)
panc8 <- NormalizeData(panc8)
panc8 <- FindVariableFeatures(panc8)
panc8 <- ScaleData(panc8)
panc8 <- RunPCA(panc8)
set.seed(2893)
panc8 <- IntegrateLayers(object = panc8, 
                         method = CoralysisIntegration, 
                         new.reduction = "integrated.coralysis",
                         batch = "replicate",
                         threads = 4)
panc8 <- RunUMAP(panc8, reduction = 'integrated.coralysis', dims = 1:30)
panc8 <- FindNeighbors(panc8, reduction = 'integrated.coralysis', dims = 1:30)
panc8 <- FindClusters(panc8)
DimPlot(panc8, group.by = c('replicate', 'ident', 'celltype'), ncol = 3) 
```
