---
title: "Running FLASH-MM on Seurat objects for differential expression analysis"
date: 'Compiled: `r format(Sys.Date(), "%B %d, %Y")`'
output:
  github_document:
    html_preview: true
    toc: true
    toc_depth: 3
  html_document:
    df_print: kable
    theme: united
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  tidy = TRUE,
  tidy.opts = list(width.cutoff = 95),
  message = FALSE,
  warning = FALSE,
  collapse = TRUE,
  comment = "#>"
)
```


This vignette demonstrates how to run FLASH-MM on Seurat objects for single-cell differential expression (DE) analysis. For demonstration purposes, we will use the interferon-beta stimulated human PBMC dataset (ifnb) to perform DE analysis within the same cell type across conditions, as described in the Seurat [differential expression testing](https://satijalab.org/seurat/articles/de_vignette) vignette.

FLASH-MM is a fast and scalable DE analysis method based on linear mixed-effects models, which help address intra-subject correlation and inter-subject variability in single-cell RNA-seq data. See [FLASHMM vignette](https://cran.r-project.org/package=FLASHMM) for details. If you find FLASH-MM useful for your publication, please cite:

> *FLASH-MM: fast and scalable single-cell differential expression analysis using linear mixed-effects models*
>
> Changjiang Xu, Delaram Pouyabahar, Veronique Voisin, Hamed Heydari & Gary D. Bader
>
> Nature Communications 17, 2026
>
> doi: [10.1038/s41467-026-69063-2](https://doi.org/10.1038/s41467-026-69063-2)
> 
> GitHub: https://github.com/BaderLab/FLASHMM


Prerequisites to install:

* [Seurat](https://satijalab.org/seurat/install)
* [SeuratWrappers](https://github.com/satijalab/seurat-wrappers)
* [SeuratData](https://github.com/satijalab/seurat-data)
* [FLASHMM](https://cran.r-project.org/package=FLASHMM)


```{r echo = TRUE, results = "hide", message = FALSE}
library(Seurat)
library(SeuratData)
library(SeuratWrappers)
library(FLASHMM)
```


## Dataset: interferon-stimulated and control PBMC

The interferon-beta stimulated human PBMCs dataset (ifnb) is available via the [SeuratData](https://github.com/satijalab/seurat-data) package.
 
```{r ifnb, echo = TRUE}
InstallData("ifnb")
ifnb <- LoadData("ifnb")
ifnb

#download donor (sample) information from the Github repo of the source data
info_repo <- "https://raw.githubusercontent.com/yelabucsf/demuxlet_paper_code/master/fig3/"
ctrl <- read.table(url(paste0(info_repo, "ye1.ctrl.8.10.sm.best")), head = T, stringsAsFactors = F)
stim <- read.table(url(paste0(info_repo, "ye2.stim.8.10.sm.best")), head = T, stringsAsFactors = F)
info <- rbind(ctrl, stim)

#rename the cell IDs by substituting the '-' into '.'
info$BARCODE <- gsub("-", ".", info$BARCODE)
#only keep the cells with high-confidence sample ID (singlets, SNG)
info <- info[grep("SNG", info$BEST), ]
#remove cells with duplicated IDs in both ctrl and stim groups
info <- info[!duplicated(info$BARCODE) & !duplicated(info$BARCODE, fromLast = T), ]

#add the sample IDs to Seurat Object ifnb
rownames(info) <- info$BARCODE
ifnb <- AddMetaData(ifnb, metadata = info[,"BEST", drop=F], col.name = "donor_id")
#remove cells without donor IDs (cells that couldn't be mapped to a donor)
ifnb <- subset(ifnb, subset = !is.na(donor_id))
dim(ifnb)

```


## Preprocessing: filtration and normalization.

The standard pre-processing workflow for scRNA-seq data in Seurat, including filtration and normalization, is described in [Seurat - Guided Clustering Tutorial](https://satijalab.org/seurat/articles/pbmc3k_tutorial).

```{r prep, echo = TRUE}
#We filter cells that have unique feature counts over 2,500 or less than 200
ifnb <- subset(ifnb, subset = nFeature_RNA > 200 & nFeature_RNA < 2500)

#We normalize the data using the log-normalization method, i.e., log(1 + 10000*counts/library.size).
ifnb <- NormalizeData(ifnb, normalization.method = "LogNormalize", scale.factor = 10000)

```


## Differential expression analysis using FLASH-MM

In this example, we are interested in identifying what genes change in different conditions for cells of the same type.

### Filtering cell types and genes 

We filter out the cell types with a low number of cells and the genes expressed very lowly.

```{r DEA-filtering}
#add log(library-size) to Seurat Object
ifnb <- AddMetaData(ifnb, metadata = log(colSums(ifnb[["RNA"]]$counts)), col.name = "loglib")

#rename "seurat_annotations" as "celltype"
colnames(ifnb@meta.data)[grep("annotations", colnames(ifnb@meta.data))] <- "celltype"

#filtering out cell types with a low number of cells
min_cells <- 100
celltype_counts <- table(ifnb$celltype)
keep_types <- names(celltype_counts)[celltype_counts >= min_cells]
ifnb <- subset(ifnb, subset = celltype %in% keep_types)

#keep genes expressed in at least 5 cells in each cell-type and treatment
celltype_trt <- paste0(ifnb@meta.data$celltype, ifnb@meta.data$stim)
gene_counts_celltypes <- do.call(cbind,
		tapply(1:ncol(ifnb), celltype_trt, 
		function(i) rowSums(ifnb[["RNA"]]$counts[, i] > 0)))
				
nCells <- 5
keep_genes <- rownames(gene_counts_celltypes)[apply(gene_counts_celltypes >= nCells, 1, all)]
ifnb <- subset(ifnb, features = keep_genes)
dim(ifnb)	

```


### Model design: fixed and random effect formulas

We define a fixed-effects model to capture systematic effects of logarithm of the library size (loglib) , cell type (celltype), and the interaction between cell type and IFN stimulation (stim), while a random-effects model is included to account for variability between donors (donor_id). The fixed and random effect variables in the model formulas cannot contain special characters such as hyphen `'-'` and empty `' '`. We rename the donor (sample) IDs by substituting `'-'` with `''`, and the cell-types by substituting `' '` with `'.'`.

```{r DEA-design}
ifnb@meta.data$donor_id <- gsub("\\-", "", ifnb@meta.data$donor_id)
ifnb@meta.data$celltype <- gsub(" ", ".", ifnb@meta.data$celltype)

#fixed-effects formula
fixed <- as.formula(~ 0 + loglib + celltype + celltype:stim)

#random-effects formulas
random <- as.formula(~ 0 + donor_id)

```


### Fitting LMMs by RunFLASHMM

We run FLASHMM on the Seurat object ifnb using the fixed and random effects we specified, with is.counts = FALSE since we are working with normalized data. 

```{r DEA-fittingLMM}

#use normalized data
t0 <- Sys.time()
fit <- RunFLASHMM(ifnb, fixed, random, is.counts = FALSE, verbose = FALSE) 
difftime(Sys.time(), t0)

#p-values
fit$p[, 1:3]
```


### Fitting LMMs with multiple random-effects components

To perform differential expression analysis between treatments (control versus interferon-beta stimulation) across all cell types, we model donor ID and cell type as random effects, as defined below.

```{r DEA-fittingLMM2}
#fixed-effects formula
fixed <- as.formula(~ loglib + stim)

#Two random-effects components
random <- list(~ 0 + donor_id, ~ 0 + celltype)

t0 <- Sys.time()
fit <- RunFLASHMM(ifnb, fixed, random, is.counts = FALSE, verbose = FALSE) 
difftime(Sys.time(), t0)

#p-values
fit$p[, 1:4]	

```
