---
title: "Running GLM-PCA on a Seurat Object"
date: 'Compiled: `r format(Sys.Date(), "%B %d, %Y")`'
output: 
  github_document:
    html_preview: false
    toc: false
  html_document:
    df_print: kable
    theme: united
---
  
This vigettte demonstrates how to run GLM-PCA, which implements a generalized version of PCA for non-normally distributed data, on a Seurat object. If you use this, please cite:

> *Feature selection and dimension reduction for single-cell RNA-Seq based on a multinomial model*
>
> F. William Townes, Stephanie C. Hicks, Martin J. Aryee & Rafael A. Irizarry 
>
> Genome Biology, 2019
>
> doi: https://doi.org/10.1186/s13059-019-1861-6
>
> GitHub: https://github.com/willtownes/glmpca
> CRAN: https://cran.r-project.org/web/packages/glmpca/index.html

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  tidy = TRUE,
  tidy.opts = list(width.cutoff = 95),
  message = FALSE,
  warning = FALSE,
  fig.height = 20,
  fig.width = 16
)
```

Prerequisites to install: 
  
* [Seurat](https://satijalab.org/seurat/install) 
* [SeuratWrappers](https://github.com/satijalab/seurat-wrappers)
* [SeuratData](https://github.com/satijalab/seurat-data) 
* [glmpca](https://github.com/willtownes/glmpca)

```{r packages}
library(Seurat)
library(SeuratData)
library(SeuratWrappers)
library(glmpca)
library(dplyr)
```

### scRNA-seq imputation on pbmc3k

To learn more about this dataset, type `?pbmc3k` 

```{r sctransform, results='hide', cache=TRUE}
InstallData("pbmc3k")
data("pbmc3k")

# Initial processing to select variable features 
pbmc3k <- NormalizeData(pbmc3k,verbose = F) %>% FindVariableFeatures(verbose=F)

# run GLM-PCA on Seurat object. 
# Uses poisson model by default
# Note that data in the counts slot is used
# We choose 10 dimensions for computational efficiency

ndims = 10
pbmc3k <- SeuratWrappers:::RunGLMPCA(pbmc3k,features = VariableFeatures(pbmc3k),L = ndims)
pbmc3k <- FindNeighbors(pbmc3k,reduction = 'glmpca',dims = 1:ndims,verbose = FALSE) %>% FindClusters(verbose = F)
pbmc3k <- RunUMAP(pbmc3k,reduction = 'glmpca',dims = 1:ndims,verbose = FALSE)
```


<!-- out.height = 8, fig.height = 20, fig.width = 16 -->

```{r explore, results='hide', cache=TRUE}

# note that we need to optimize these visualizations

# visualize original and imputed values
features.plot <- c('CD3D', 'MS4A1', 'CD8A', 'GZMK', 'GZMB', 'FCGR3A')
DimPlot(pbmc3k)
FeaturePlot(pbmc3k, features.plot, ncol = 2)
```

