---
title: "Running LocalMAP on a Seurat Object"
date: 'Compiled: `r format(Sys.Date(), "%B %d, %Y")`'
output:
  github_document:
    html_preview: false
    toc: false
  html_document:
    toc: false
    df_print: paged
---

This vignette demonstrates how to run **LocalMAP** on a Seurat object. LocalMAP is a
variant of PaCMAP that adds a local graph-adjustment stage in phase 3: further-pair
partners are resampled to points that are already close in the low-dimensional
embedding (within `low_dist_thres`), and the nearest-neighbour gradient term is
scaled by `low_dist_thres / (2 * sqrt(d_ij))`. Compared with base PaCMAP this
tightens local structure while retaining the same global scaffold.

If you use LocalMAP, please cite:

> *Dimension Reduction with Locally Adjusted Graphs*
>
> Yingfan Wang, Yiyang Sun, Haiyang Huang and Cynthia Rudin
>
> Preprint, 2024
>
> GitHub: https://github.com/YingfanWang/PaCMAP

and the original PaCMAP references linked from the [PaCMAP vignette](pacmap.Rmd).

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  fig.width = 10
)
```

## Prerequisites

* [Seurat](https://satijalab.org/seurat/install)
* [SeuratWrappers](https://github.com/satijalab/seurat-wrappers)
* [SeuratData](https://github.com/satijalab/seurat-data)
* [pacmapr](https://github.com/williamsyy/pacmap-for-R) — native R + Rcpp implementation of PaCMAP and LocalMAP.

To install `pacmapr` directly:

```{r install_pacmapr, eval=FALSE}
# install.packages("remotes")
remotes::install_github("williamsyy/pacmap-for-R", subdir = "pacmapr")
```

No Python, conda, or `reticulate` installation is required.

```{r packages}
library(Seurat)
library(SeuratData)
library(SeuratWrappers)
```

### LocalMAP on PBMC3k

```{r localmap, cache=TRUE, cache.lazy=TRUE}
InstallData("pbmc3k")
pbmc3k.final <- LoadData("pbmc3k", type = "pbmc3k.final")

pbmc3k.final <- UpdateSeuratObject(pbmc3k.final)
pbmc3k.final <- FindVariableFeatures(pbmc3k.final)

# Run LocalMAP on the Seurat object.
pbmc3k.final <- RunLocalMAP(
  object   = pbmc3k.final,
  features = VariableFeatures(pbmc3k.final)
)
```

```{r explore, fig.width=6}
features.plot <- c("CD3D", "MS4A1", "CD8A", "GZMK", "GZMB", "FCGR3A")
DimPlot(object = pbmc3k.final, reduction = "localmap")
```

```{r explore2, fig.height=10}
pbmc3k.final <- NormalizeData(pbmc3k.final, verbose = FALSE)
FeaturePlot(pbmc3k.final, features.plot, ncol = 2, reduction = "localmap")
```

You can also run LocalMAP on the dims of an existing reduction:

```{r localmap_dim, cache=TRUE, cache.lazy=TRUE}
pbmc3k.final <- RunLocalMAP(object = pbmc3k.final, dims = 2:5)
DimPlot(object = pbmc3k.final, reduction = "localmap")
```

### Tuning `low_dist_thres`

`low_dist_thres` (default 10, per Wang et al.) is the LocalMAP-specific knob. It
controls (a) the low-D distance below which further-pair partners are resampled in
phase 3, and (b) the strength of the NN-gradient scaling. Smaller values push the
embedding to tighten local clusters more aggressively; larger values behave more
like base PaCMAP.

```{r localmap_thres, eval=FALSE}
tight <- RunLocalMAP(
  object         = pbmc3k.final,
  features       = VariableFeatures(pbmc3k.final),
  low_dist_thres = 3,
  reduction.name = "localmap_tight"
)
loose <- RunLocalMAP(
  object         = pbmc3k.final,
  features       = VariableFeatures(pbmc3k.final),
  low_dist_thres = 30,
  reduction.name = "localmap_loose"
)
```

### Notes on defaults

Wrapper defaults match the reference implementation:

* `num_iters = 250L` — expands to `c(100, 100, 250)` = 450 total iterations.
* `n.neighbors = NULL` — auto-selected: 10 for `n ≤ 10 000`, otherwise `round(10 + 15 * (log10(n) - 4))`.
* `distance_method = "euclidean"`, `MN_ratio = 0.5`, `FP_ratio = 2`.
* `apply_pca = TRUE`, `init = "random"`.
* `low_dist_thres = 10`.
* `n_threads = NULL` — uses `parallel::detectCores() - 1L`.
