---
title: "Calculating Trajectories with Monocle 3 and Seurat"
output:
  html_document:
    df_print: kable
    theme: united
  github_document:
    html_preview: no
    toc: no
date: 'Compiled: `r format(Sys.Date(), "%B %d, %Y")`'
---
  
This vigettte demonstrates how to run trajectory inference and pseudotime calculations with Monocle 3 on Seurat objects. If you use Monocle 3, please cite:

> *The single-cell transcriptional landscape of mammalian organogenesis*
>
> Junyue Cao, Malte Spielmann, Xiaojie Qiu, Xingfan Huang, Daniel M. Ibrahim, Andrew J. Hill, Fan Zhang, Stefan Mundlos, Lena Christiansen, Frank J. Steemers, Cole Trapnell & Jay Shendure
>
> doi: [10.1038/s41586-019-0969-x](https://doi.org/10.1038/s41586-019-0969-x)
>
> Website: https://cole-trapnell-lab.github.io/monocle3/

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  tidy = TRUE,
  tidy.opts = list(width.cutoff = 95),
  message = FALSE,
  warning = FALSE,
  fig.height = 20,
  fig.width = 16
)
```

Prerequisites to install:

* [Seurat](https://satijalab.org/seurat/install)
* [SeuratWrappers](https://github.com/satijalab/seurat-wrappers)
* [SeuratData](https://github.com/satijalab/seurat-data)
* [Monocle 3](https://cole-trapnell-lab.github.io/monocle3/docs/installation/)

```{r install_deps, echo=FALSE}
if (!requireNamespace("monocle3", quietly = TRUE)) {
  setRepositories(ind = 1:2)
  remotes::install_github(
    repo = "cole-trapnell-lab/monocle3",
    upgrade = FALSE
  )
}
```

```{r memsafe, echo=FALSE, results="hide"}
gc(verbose = FALSE)
options(Seurat.memsafe = TRUE)
```

```{r packages}
library(monocle3)
library(Seurat)
library(SeuratData)
library(SeuratWrappers)
library(ggplot2)
library(patchwork)
library(magrittr)
```

## HCA Bone Marrow 40k

```{r load_hca}
InstallData("hcabm40k")
data("hcabm40k")
```

```{r preprocess_hca, results='hide'}
hcabm40k <- SplitObject(hcabm40k, split.by = "orig.ident")
for (i in seq_along(hcabm40k)) {
  hcabm40k[[i]] <- NormalizeData(hcabm40k[[i]]) %>% FindVariableFeatures()
}
```

```{r gc1, echo=FALSE, results="hide"}
gc(verbose = FALSE)
```

```{r normalize_part_two, results="hide"}
features <- SelectIntegrationFeatures(hcabm40k)
for (i in seq_along(along.with = hcabm40k)) {
  hcabm40k[[i]] <- ScaleData(hcabm40k[[i]], features = features) %>%
    RunPCA(features = features)
}
```

```{r gc2, echo=FALSE, results="hide"}
gc(verbose = FALSE)
```

```{r anchors, results='hide'}
anchors <- FindIntegrationAnchors(
  hcabm40k,
  reference = c(1, 2),
  reduction = "rpca",
  dims = 1:30
)
```

```{r gc3, echo=FALSE, results="hide"}
rm(hcabm40k)
gc(verbose = FALSE)
```

```{r integration, results="hide"}
integrated <- IntegrateData(anchors, dims = 1:30)
```

```{r gc4, echo=FALSE, results="hide"}
rm(anchors)
gc(verbose = FALSE)
```

```{r dimreduc_hca, results='hide', fig.height=10, fig.width=16}
integrated <- ScaleData(integrated)
integrated <- RunPCA(integrated)
integrated <- RunUMAP(integrated, dims = 1:30, reduction.name = "UMAP")
integrated <- FindNeighbors(integrated, dims = 1:30)
integrated <- FindClusters(integrated)
DimPlot(integrated, group.by = c("orig.ident", "ident"))
```

```{r gc5, echo=FALSE, results="hide"}
gc(verbose = FALSE)
```

```{r monocle3_clustering, results='hide', fig.height=10, fig.width=16}
cds <- as.cell_data_set(integrated)
cds <- cluster_cells(cds)
p1 <- plot_cells(cds, show_trajectory_graph = FALSE)
p2 <- plot_cells(cds, color_cells_by = "partition", show_trajectory_graph = FALSE)
wrap_plots(p1, p2)
```

```{r gc6, echo=FALSE, results="hide"}
rm(integrated)
gc(verbose = FALSE)
```

```{r subset_partitions, results="hide", fig.height=10, fig.width=16}
integrated.sub <- subset(as.Seurat(cds), monocle3_partitions == 1)
cds <- as.cell_data_set(integrated.sub)
cds <- learn_graph(cds)
plot_cells(
  cds,
  label_groups_by_cluster = FALSE,
  label_leaves = FALSE,
  label_branch_points = FALSE
)
```

```{r gc7, echo=FALSE, results="hide"}
gc(verbose = FALSE)
```

```{r pseudotime, results="hide", fig.height=10, fig.width=16}
max.avp <- which.max(unlist(FetchData(integrated.sub, "AVP")))
max.avp <- colnames(integrated.sub)[max.avp]
cds <- order_cells(cds, root_cells = max.avp)
plot_cells(
  cds,
  color_cells_by = "pseudotime",
  label_cell_groups = FALSE,
  label_leaves = FALSE,
  label_branch_points = FALSE
)
integrated.sub <- as.Seurat(cds)
FeaturePlot(integrated.sub, "monocle3_pseudotime")
```
