---
title: "Running PaCMAP on a Seurat Object"
date: 'Compiled: `r format(Sys.Date(), "%B %d, %Y")`'
output:
  html_document:
    toc: false
    df_print: paged
  github_document:
    html_preview: false
    toc: false
---
  
This vignette demonstrates how to run PaCMAP, a dimensionality reduction method that can
be used for providing robust and trustworthy visualization, on a Seurat object.
If you use our work, please cite both papers:

> *Understanding How Dimension Reduction Tools Work: An Empirical Approach to Deciphering t-SNE, UMAP, TriMap, and PaCMAP for Data Visualization*
>
> Yingfan Wang, Haiyang Huang, Cynthia Rudin & Yaron Shaposhnik
>
> Journal of Machine Learning Research, 2021
>
> doi: https://doi.org/10.48550/arXiv.2012.04456
> 
> *Towards a comprehensive evaluation of dimension reduction methods for transcriptomic data visualization*
>
> Haiyang Huang, Yingfan Wang, Cynthia Rudin and Edward P. Browne
>
> Communications biology, 2022
>
> doi: https://doi.org/10.1038/s42003-022-03628-x
>
> GitHub: https://github.com/YingfanWang/PaCMAP

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  fig.width = 10
)
```

Prerequisites to install: 
  
* [Seurat](https://satijalab.org/seurat/install) 
* [SeuratWrappers](https://github.com/satijalab/seurat-wrappers)
* [SeuratData](https://github.com/satijalab/seurat-data)
* [Reticulate](https://rstudio.github.io/reticulate/)

In addition to R packages, PaCMAP relies on Python to deliver high performance. 
To streamline the installation process and make environment management easier, 
we strongly recommend you to use anaconda(https://www.anaconda.com/download) or
miniconda(https://docs.anaconda.com/miniconda/miniconda-install/) for managing Python
environments. Below, we provide step-by-step instructions on how to properly install PaCMAP 
**after** you have installed one of these tools.

Create a conda environment with PaCMAP installed:

```{bash conda environment setup, eval=FALSE}
conda create -n "pacmap" python=3.12  # Install in the environment called "pacmap"
conda activate pacmap
conda install -y conda-forge::pacmap
```

To run PaCMAP, you need to connect your R console to the corresponding conda environment.
If your Conda/Miniconda installation is located in a non-default directory, you might
set up the conda variable as `/path/to/your/conda`. This ensures the correct environment
is used during the installation.

```{r conda environment connects, eval=FALSE}
reticulate::use_condaenv(condaenv = "pacmap", conda = "auto")
```

```{r packages}
library(Seurat)
library(SeuratData)
library(SeuratWrappers)
```

### PaCMAP on PBMC3k

To learn more about this dataset, type `?pbmc3k` 

```{r pacmap, cache=TRUE, cache.lazy=TRUE}
InstallData("pbmc3k")
pbmc3k.final <- LoadData("pbmc3k",type="pbmc3k.final")

# Initial processing to select variable features
pbmc3k.final <- UpdateSeuratObject(pbmc3k.final)
pbmc3k.final <- FindVariableFeatures(pbmc3k.final)

# run PaCMAP on Seurat object. 
pbmc3k.final <- RunPaCMAP(object = pbmc3k.final, features=VariableFeatures(pbmc3k.final))
```

```{r explore, fig.width=6}
# visualize markers
features.plot <- c('CD3D', 'MS4A1', 'CD8A', 'GZMK', 'GZMB', 'FCGR3A')
DimPlot(object=pbmc3k.final,reduction="pacmap")
```

```{r explore2, fig.height=10}
pbmc3k.final <- NormalizeData(pbmc3k.final, verbose = FALSE) 
FeaturePlot(pbmc3k.final, features.plot, ncol = 2, reduction="pacmap")
```
