---
title: "Running PaCMAP on a Seurat Object"
date: 'Compiled: `r format(Sys.Date(), "%B %d, %Y")`'
output:
  github_document:
    html_preview: false
    toc: false
  html_document:
    toc: false
    df_print: paged
---

This vignette demonstrates how to run PaCMAP, a dimensionality reduction method that
provides robust and trustworthy visualization, on a Seurat object. If you use our work,
please cite:

> *Understanding How Dimension Reduction Tools Work: An Empirical Approach to Deciphering t-SNE, UMAP, TriMap, and PaCMAP for Data Visualization*
>
> Yingfan Wang, Haiyang Huang, Cynthia Rudin & Yaron Shaposhnik
>
> Journal of Machine Learning Research, 2021
>
> doi: https://doi.org/10.48550/arXiv.2012.04456
>
> *Towards a comprehensive evaluation of dimension reduction methods for transcriptomic data visualization*
>
> Haiyang Huang, Yingfan Wang, Cynthia Rudin and Edward P. Browne
>
> Communications biology, 2022
>
> doi: https://doi.org/10.1038/s42003-022-03628-x
>
> GitHub: https://github.com/YingfanWang/PaCMAP

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  fig.width = 10
)
```

## Prerequisites

* [Seurat](https://satijalab.org/seurat/install)
* [SeuratWrappers](https://github.com/satijalab/seurat-wrappers)
* [SeuratData](https://github.com/satijalab/seurat-data)
* [pacmapr](https://github.com/williamsyy/pacmap-for-R) — native R + Rcpp implementation of PaCMAP.

`SeuratWrappers` lists `pacmapr` under `Remotes:`, so installing SeuratWrappers via
`remotes::install_github("satijalab/seurat-wrappers")` pulls it in automatically.
To install it directly:

```{r install_pacmapr, eval=FALSE}
# install.packages("remotes")
remotes::install_github("williamsyy/pacmap-for-R", subdir = "pacmapr")
```

No Python, conda, or `reticulate` installation is required — everything runs natively
in R via C++.

```{r packages}
library(Seurat)
library(SeuratData)
library(SeuratWrappers)
```

### PaCMAP on PBMC3k

To learn more about this dataset, type `?pbmc3k`.

```{r pacmap, cache=TRUE, cache.lazy=TRUE}
InstallData("pbmc3k")
pbmc3k.final <- LoadData("pbmc3k", type = "pbmc3k.final")

# Initial processing to select variable features
pbmc3k.final <- UpdateSeuratObject(pbmc3k.final)
pbmc3k.final <- FindVariableFeatures(pbmc3k.final)

# Run PaCMAP on the Seurat object (native R backend via pacmapr).
pbmc3k.final <- RunPaCMAP(
  object   = pbmc3k.final,
  features = VariableFeatures(pbmc3k.final)
)
```

```{r explore, fig.width=6}
features.plot <- c("CD3D", "MS4A1", "CD8A", "GZMK", "GZMB", "FCGR3A")
DimPlot(object = pbmc3k.final, reduction = "pacmap")
```

```{r explore2, fig.height=10}
pbmc3k.final <- NormalizeData(pbmc3k.final, verbose = FALSE)
FeaturePlot(pbmc3k.final, features.plot, ncol = 2, reduction = "pacmap")
```

You can also specify dims of an existing reduction (e.g. PCA) as input to PaCMAP:

```{r pacmap_dim, cache=TRUE, cache.lazy=TRUE}
pbmc3k.final <- RunPaCMAP(object = pbmc3k.final, dims = 2:5)
DimPlot(object = pbmc3k.final, reduction = "pacmap")
```

### Notes on defaults

The wrapper defaults mirror the reference Python PaCMAP:

* `num_iters = 250L` — expands inside `pacmapr` to `c(100, 100, 250)` = 450 total optimisation iterations.
* `n.neighbors = NULL` — auto-selected: 10 for `n <= 10 000`, otherwise `round(10 + 15 * (log10(n) - 4))`.
* `distance_method = "euclidean"`, `MN_ratio = 0.5`, `FP_ratio = 2`.
* `apply_pca = TRUE` — PaCMAP applies PCA-to-100 preprocessing when the input has more than 100 features.
* `init = "random"`; pass `"pca"` or a numeric initial-embedding matrix if you prefer.
* `n_threads = NULL` — uses `parallel::detectCores() - 1L`. Set to `1L` for a fully single-threaded run.
